Appendix S1. Linear regression models and non-linear population dynamics

### 2024-11-26 manuscript/Appendix_S1_v02.Rmd
Updated S1, knit.

# One species, logistic growth
Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.

@@

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- 1:t
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
    linetype = "dashed", color = "gray") +
  theme_classic()
```

@@

# One species, logistic growth, environmental covariate
We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.

@@

# Two species, logistic growth, competition
Here we investigate how a second species impacts the inference we can make from linear models.

@@

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs12}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
```
## Linear statistical model with covariate for both species

@@

```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 2 in the previous time step $N_t$."}
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
colnames(df) <- c("Nt1","N1","N2")
m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)

cor(df$Nt1, predict(m.3.lm))
summary(m.3.lm)
xx <- seq(0,100, length=1000)
prediction<-data.frame(N2=xx,N1=100)
plot(df$N2, df$Nt1, xlab = "N2", ylab="N.t+1" , pch=16)
lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)
```

```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$. Fitted (line) and observed (points) values are shown.", warning=FALSE}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
# ggplot(data = df, aes(x = N2, y = Nt1)) +
#   geom_point() +
#   geom_smooth(method = "lm", formula = y ~ x) + theme_classic()
```

@@

```{r}
# prepare data in HMSC format
Y <- as.matrix(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
XData <- df[,2:3]
studyDesign = data.frame(sample = as.factor(1:(t-1)))
studyDesign$sample <- as.factor(studyDesign$sample)
rL.sample = HmscRandomLevel(units = studyDesign$sample)
m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

The model fixed effect estimates match well with those from the AR1 and linear model.

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.3.ar.n1$coef
m.3.ar.n1$coef[2]*(1-m.3.ar.n1$coef[1])
# linear model
summary(m.3.lm.n1)$coefficients[1:3,1:2]
# Bayesian estimates
summary(m3.post.hmsc$Beta)$statistics[1:2,1:2]
```

```{r figs18, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs18} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions."}
preds = computePredictedValues(m.3.sample)
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
colnames(df) <- c("Nt1","Nt2","N1","N2")
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```

```{r}
knitr::knit_exit()
```

@@

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.7
r2.0 <- 1.7
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") = theme_classic()
```

@@

```{r}
# explanatory power
preds = computePredictedValues(m.4.sample)
#evaluateModelFit(hM = m.4.sample, predY = preds)
# predictive power / cross-validation
#partition = createPartition(m.4.sample, nfolds = 2)
#preds = computePredictedValues(m.4.sample, partition = partition, nParallel = nChains)
#evaluateModelFit(hM = m.4.sample, predY = preds)
## xx
#preds = computePredictedValues(m.4.sample, partition=partition, partition.sp=c(1,2), mcmcStep=10, nParallel = nChains)
#evaluateModelFit(hM=m.4.sample, predY=preds)
# xx
#etaPost=getPostEstimate(m.4.sample, "Eta")
#lambdaPost=getPostEstimate(m.4.sample, "Lambda")
#biPlot(m.4.sample, etaPost = etaPost, lambdaPost = lambdaPost, factors = c(1,2),"X1")
```
```{r}
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="blue")
lines(df$n1_975,col="blue")
lines(df$n2_25,col="blue")
lines(df$n2_975,col="blue")
```

The residual associations should not be interpreted as interaction coefficients, but rather used for prediction. We can see here they are effective for this (shown above is the observed and model-predicted 95% CI).

# Species growth, competition, with environmental change

@@

In this example, the local environmental optimum trait value $E = 0.8$, and Species 1 $x_1 = 0.6$, while $x_1 = 0.8$.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.7
r2.0 <- 1.7
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
E.0 <- 0.8
x1.0 <- 0.6
x2.0 <- 0.8

# model function
disc_LV_E <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1,x2) {
    Nt1 <- ((r1*exp(-(E-x1)^2))*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- ((r2*exp(-(E-x2)^2))*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  res <- disc_LV_E(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x1.0,x2=x2.0)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
```{r}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(log(N$N1),log(N$N2)))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t]))
colnames(df) <- c("Nt1","Nt2")
# prepare data in HMSC format
Y <- as.matrix(cbind(df$Nt1,df$Nt2))
XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2)
colnames(XData) <- c("n1","n2","E","Esq")
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

postBeta = getPostEstimate(m.5.sample, parName = "Beta")
plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```
We can see that the Environment has an overall negative impact for both species as $E$ drifts randomly towards more positive values (where Species 2 is closer because it has a higher trait value). We also see that the residual species association effects are strongly positive.

@@

```{r warning=FALSE}
VP <- computeVariancePartitioning(m.5.sample,group=c(1,1,1),groupnames="Env")
plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
```
We evaluate this by greatly increasing the effects of interspecific competition (while decreasing the strength of environmental variation).

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.7
r2.0 <- 1.7
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.015
alpha.21 <- 0.02
E.0 <- 0.8
x1.0 <- 0.8
x2.0 <- 0.8

>>>>

# Model A: One species, logistic growth
Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.

@@

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 10, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- 1:t
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
    linetype = "dashed", color = "gray") +
  theme_classic()
```

@@

# Model B: One species, logistic growth, environmental covariate
We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.

@@

# Model C: Two species, logistic growth, competition
Here we investigate how a second species impacts the inference we can make from linear models.

@@

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$. The simulation is generated using the `disc_LV_comp` function in the `ecoevor` package.

```{r}
# Initial conditions
N0 <- c(10,10)
r <- c(1.67,1.7)
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
# make sure the interaction matrix is correct
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N[1,] <- N0
for (i in 2:t) {
  N[i,] <- disc_LV_comp(r=r,N0=N[i-1,],alpha=alpha)
}
```
```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs12}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
```
## Linear statistical model with covariate for both species

@@

```{r}
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
colnames(df) <- c("Nt1","N1","N2")
m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)
cor(df$Nt1, predict(m.3.lm))
summary(m.3.lm)
```

```{r figs16, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs16}Population size (logarithm) in species 1 at one time step $N_{1,t+1}$ as a function of log-population size of Species 2 in the previous time step $N_t$."}
xx <- seq(0,100, length=1000)
prediction<-data.frame(N2=xx,N1=100)
plot(df$N2, df$Nt1, xlab = "N2", ylab="N.t+1" , pch=16)
lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)
```

```{r figs17, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs17} Linear model-predicted population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$. Fitted (line) and observed (points) values are shown.", warning=FALSE, message=FALSE}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
# ggplot(data = df, aes(x = N2, y = Nt1)) +
#   geom_point() +
#   geom_smooth(method = "lm", formula = y ~ x) + theme_classic()
```

@@

```{r}
# prepare data in HMSC format
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
XData <- data.frame(dat[1:(t-1),-3])
colnames(XData) <- c(paste0("n",1:length(r)))
studyDesign = data.frame(sample = as.factor(1:(t-1)))
studyDesign$sample <- as.factor(studyDesign$sample)
rL.sample = HmscRandomLevel(units = studyDesign$sample)
m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL.sample))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

The direction of HMSC model fixed effect estimates matches well with those from the AR1 and linear model.

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.3.ar.n1$coef
m.3.ar.n1$coef[2]*(1-m.3.ar.n1$coef[1])
# linear model
summary(m.3.lm.n1)$coefficients[1:3,1:2]
# Bayesian estimates
summary(m3.post.hmsc$Beta)$statistics[1:2,1:2]
```
The HMSC model correctly shows the negative association between the species (shown here as correlations - JSDM CHapter 7 tells us the intra-specific correlation is always 1).

```{r figs18, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs18} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions. Lines represent the 95% highest probability density interval (HPDI) for predictions at each time point."}
preds = computePredictedValues(m.3.sample)
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$N1[1:(t-1)],dat$N2[1:(t-1)]))
colnames(df) <- c("Nt1","Nt2","N1","N2")
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```

@@

```{r}
# Initial conditions
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE) # careful to make sure you have correct interaction matrix
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N[1,] <- N0
for (i in 2:t) {
  N[i,] <- disc_LV_comp(r=r,N0=N[i-1,],alpha=alpha)
}
```
```{r figs19, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs19} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic()
```

@@

```{r}
# explanatory power
preds = computePredictedValues(m.4.sample)
```
```{r figs20, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs20} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions. Lines represent the 95% HPDI for predictions at each time point."}
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$Nt1,pch=19,col="black",ylim=c(0,5))
points(df$Nt2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```

The residual associations should not be interpreted as interaction coefficients, but rather used for prediction. We can see here they are effective for this (shown above is the observed and model-predicted 95% highest probability density interval - HPDI).

# Model D: Species growth, competition, with environmental change

@@

In this example, the local environmental optimum trait value $E = 0.8$, and Species 1 $x_1 = 0.6$, while $x_2 = 0.8$. The simulation is generated using the `disc_LV_E` function in the `ecoevor` package.

```{r}
# Initial conditions
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
E.0 <- 0.8
x <- c(0.6,0.8)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
  E[i] <- E[i-1] + rnorm(1,0,0.1)
}
```
```{r figs21, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs21} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.005, and $\\alpha_{21}$ = 0.01."}
# Plot simulation: ggplot
gdat <- N
gdat$time <- 1:t
dat <- melt(gdat, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r[1] - 1)/alpha.11),linetype = "dashed", color = "gray") + theme_classic() + ylab("N")
```
```{r}
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
# prepare data in HMSC format
XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.5.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.5.sample <- sampleMcmc(m.5.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

@@

postBeta = getPostEstimate(m.5.sample, parName = "Beta")
plotBeta(m.5.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```
We can see that the Environment has an overall negative impact for Species 1 as $E$ drifts randomly towards more positive values (and thus further from Species 1 trait value). We also see that the residual species association for species 1 and 2 are negative.

@@

```{r warning=FALSE}
VP <- computeVariancePartitioning(m.5.sample,group=c(1,1,1),groupnames="Env")
plotVariancePartitioning(m.5.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
```
We evaluate whether variation partition correctly detects the relative importance of environment and species interactions by greatly increasing the effects of interspecific competition and decreasing the strength of environmental variation.

```{r}
# Simulate initial species population growth with environment fluctuations
N0 <- c(10,10)
r <- c(1.7,1.7)
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.015
alpha.21 <- 0.02
alpha <- matrix(c(alpha.11,alpha.21,alpha.12,alpha.22),nrow=2,byrow=FALSE)
E.0 <- 0.8
x <- c(0.8,0.8)
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i,] <- disc_LV_E(r=r,N0=N[i-1,],alpha=alpha,E=E[i-1],x=x)
  E[i] <- E[i-1] + rnorm(1,0,0.001)
}
```
```{r figs22, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs22} Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model with competition and a changing environment, with parameters $r_1 = r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.01, $\\alpha_{12}$ = 0.015, and $\\alpha_{21}$ = 0.02."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + ylab("N") + theme_classic()
```
```{r}
## prepare data in HMSC format
dat <- as.data.frame(log(N))
dat$time <- 1:t
df <- data.frame(dat[(2:t),-3])
Y <- as.matrix(df)
XData <- data.frame(dat[1:(t-1),-3],E[1:(t-1)],E[1:(t-1)]^2)
colnames(XData) <- c(paste0("n",1:length(r)),"E","Esq")
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.6.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 100*thin
verbose <- 500*thin
# sample MCMC
m.6.sample <- sampleMcmc(m.6.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```
```{r}
OmegaCor = computeAssociations(m.6.sample)
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.6.sample$rLNames[1]), mar=c(0,0,1,0))
OmegaCor
```
```{r}
Gradient <- constructGradient(m.6.sample,focalVariable="E",non.focalVariables=list(Esq=list(2)),ngrid=39)
predY <- predict(m.6.sample,XData=Gradient$XDataNew,expected=TRUE)
a <- plotGradient(m.6.sample,Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="E_t",ylab="predicted N1_t+1")
b <- plotGradient(m.6.sample,Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="E_t",ylab="predicted N2_t+1")

postBeta = getPostEstimate(m.6.sample, parName = "Beta")
plotBeta(m.6.sample, post = postBeta, param = "Mean", supportLevel = 0.7)
```
With the reduced impact of the environment, we now see the strong negative associations between the two competing species. We also correctly see a relatively weak impact of environment for Species 1 abundance and a stronger impact for Species 2.

For both species, the relative importance of environment has decreased relative to the species association effects.

```{r warning=FALSE}
VP <- computeVariancePartitioning(m.6.sample,group=c(1,1,1),groupnames="Env")
plotVariancePartitioning(m.6.sample,VP,args.legend=list(cex=0.75,bg="transparent"))
```
```{r figs23, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs23} Observed (points) and HMSC model-predicted population size (logarithm) for Species 1 (black) and Species 2 (orange), using environment $E_t$ as a fixed effect predictor and specifying temporal random effects to capture the species interactions. Lines represent the 95% highest probability density interval (HPDI) for predictions at each time point."}
preds = computePredictedValues(m.6.sample)
df$n1_25 <- NA
df$n1_975 <- NA
df$n2_25 <- NA
df$n2_975 <- NA
for(i in 1:39){
  red <- preds[i,,]
  a <- apply(red,1,quantile,probs=c(.025,.975))
  df$n1_25[i] <- a[1,1]
  df$n1_975[i] <- a[2,1]
  df$n2_25[i] <- a[1,2]
  df$n2_975[i] <- a[2,2]
}
plot(df$N1,pch=19,col="black",ylim=c(-1,5),ylab="N")
points(df$N2,pch=19,col="orange")
lines(df$n1_25,col="black")
lines(df$n1_975,col="black")
lines(df$n2_25,col="orange")
lines(df$n2_975,col="orange")
```
# Model E: Species growth, competition, environmental change, and trait evolution
